---
title: "Bayesian Z-residual Diagnostics for Nonlinear Covariate Effects In Hurdle NB Models (Section 3.3 and S1.2)"
subtitle: "A hurdle negative binomial regression example"
author: "Anonymous Author"
date: today
format:
  html:
    toc: true
    number-sections: true
    page-layout: full
    code-fold: true
    html-math-method: mathjax
execute:
  echo: true
  warning: false
  message: false
  cache: true
  eval: true
vignette: >
  %\VignetteIndexEntry{Bayesian Z-residual diagnostics for nonlinear covariate effects in Hurdle NB models}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

<!-- VERSION: hnb-demo-v7-split-hnb-nb-figures; one true model, three misspecified models, separate HNB/NB figures without LOWESS curves, replicated Z-residual p-value histograms, and one summary table. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)

set.seed(123)
```

## Installing and Loading Required Packages

**Install Zresidual**

```{r install-zresidual}
#| eval: false
#| output: false
#| message: false

#During package development, run this once from the package root before rendering:
#renv::install(here::here(), dependencies = TRUE)

#For a clean external environment, install the released or review version instead:
#if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
#remotes::install_github("anonymous/Zresidual", upgrade = "never", dependencies = TRUE)
```

**Loading required packages**

```{r load-packages}
#| output: false
#| message: false
#| warning: false

pkgs <- c("brms", "dplyr", "tibble", "gt", "loo", "Zresidual")
missing_pkgs <- pkgs[!vapply(pkgs, requireNamespace, logical(1), quietly = TRUE)]

if (length(missing_pkgs)) {
  stop(
    "Please install the following packages before rendering this vignette: ",
    paste(missing_pkgs, collapse = ", "),
    call. = FALSE
  )
}

invisible(lapply(pkgs, function(p) {
  suppressPackageStartupMessages(library(p, character.only = TRUE))
}))
```

## Simulation and Fitting Settings

This demo uses the fixed representative dataset and fitted Bayesian models distributed with the vignette. These files are stored under `vignettes/articles/resources/demo_brms_HNB_nl` so that the rendered example uses the same data as the manuscript rather than a user-specific local directory.

The goal is not to reproduce a full simulation study. Instead, the file shows a compact workflow: load the paper dataset, load the saved Bayesian model fits, compute posterior and ISCV Z-residuals, compare them with PPC and HPC diagnostics, and summarize the resulting diagnostic p-values.

```{r simulation-settings}
# Fixed sample size used by the paper dataset distributed with this vignette.
n_sample <- 50

sample_tag <- paste0("N", n_sample)
file_tag <- paste0("_n", n_sample)

# One representative dataset only. This is not a 500-replication run.
representative_id <- 1L
force_recalc <- FALSE
force_refit <- force_recalc
force_rerun <- force_recalc
force_recompute_zres <- force_recalc
force_recompute_ppc_hpc <- force_recalc

# Holdout proportion for the single-dataset HPC calculation.
test_prop <- 0.20

# Data-generating parameters from the simulation code.
beta0 <- -5
beta1 <- c(0.6, -5)

alpha0 <- -1
alpha1 <- c(6, 1)

true_size <- 4

x1_min <- 0.5
x1_max <- 15


brms_backend <- "rstan"
brms_cores <- 1
brms_chains <- 4
brms_iter <- 2000
brms_warmup <- 1000

ndraws_use <- 2000
seed_use <- 20260511
x_check <- c("lp", "x1")

# Number of auxiliary randomizations for a single discrete-response dataset.
# This is not the number of simulated datasets.
nrep_z <- 500

options(mc.cores = brms_cores)

if (requireNamespace("rstan", quietly = TRUE)) {
  rstan::rstan_options(auto_write = TRUE)
}

make_k_anova <- function(n_test) {
  as.integer(max(3L, min(10L, floor(n_test / 5))))
}

hnb_family <- brms::hurdle_negbinomial(
  link = "log",
  link_shape = "identity",
  link_hu = "logit"
)

nb_family <- brms::negbinomial(
  link = "log",
  link_shape = "identity"
)

required_cols <- c(
  "y", "x1", "x2",
  "z1", "z2", "z3", "z4", "z5",
  "zeros"
)

get_vignette_dir <- function() {
  input <- knitr::current_input()
  if (!is.null(input) && length(input) == 1L && nzchar(input)) {
    return(dirname(normalizePath(input, mustWork = FALSE)))
  }

  # Fallback for interactive sourcing.
  getwd()
}

vignette_dir <- get_vignette_dir()

# The reviewer-facing archive stores the fixed paper dataset and saved fits here.
resources_path <- file.path(vignette_dir, "resources", "demo_brms_HNB_nl")

data_file <- file.path(resources_path, paste0("simulated_hnb_data_N", n_sample, ".rds"))
fits_full_file <- file.path(resources_path, paste0("fits_full_all_N", n_sample, ".rds"))
fits_hpc_file <- file.path(resources_path, paste0("fits_hpc_all_N", n_sample, ".rds"))

if (!dir.exists(resources_path)) {
  stop(
    "Cannot find the resources directory: ", resources_path,
    "\nPlease keep `resources/demo_brms_HNB_nl` next to this vignette file.",
    call. = FALSE
  )
}

if (!file.exists(data_file)) {
  stop(
    "Cannot find the fixed paper dataset: ", data_file,
    "\nExpected file name: simulated_hnb_data_N", n_sample, ".rds",
    call. = FALSE
  )
}

if (!file.exists(fits_full_file)) {
  stop(
    "Cannot find the saved full-data model fits: ", fits_full_file,
    "\nExpected file name: fits_full_all_N", n_sample, ".rds",
    call. = FALSE
  )
}
```

## Diagnostic Goal and Candidate Models

This example uses data generated from a logarithmic hurdle negative binomial (HNB) model and then fits four candidate Bayesian models to the same dataset. The first model is the data-generating model, while the other three deliberately introduce different types of misspecification:

1. **HNB-log**: true response family and true logarithmic functional form.
2. **HNB-linear**: true response family but wrong linear functional form for `x1`.
3. **NB-log**: wrong response family. Although the count mean uses a logarithmic term, the omitted hurdle zero/count mechanism can be projected into the NB mean structure, so the effective functional form can also be misspecified.
4. **NB-linear**: wrong response family and wrong linear functional form for `x1`.

The purpose is to show how different diagnostics respond to different types of model deficiency. Overall checks such as Z-SW assess whether the residual distribution is close to standard normal, while the targeted Z-AOV test against `x1` checks whether the residual mean varies systematically across the covariate. PPC and HPC are included as predictive-check comparisons based on analogous residual summaries.

Because the response is discrete, RSP Z-residuals use auxiliary randomization. In this demo, each fitted model is evaluated using `nrep_z = 500` repeated auxiliary randomizations. The replicated Z-residual p-values are shown as histograms and summarized by the distribution-free upper-bound p-value.

## Data-Generating Model

For each observation, generate

$$
X_{i,1}\sim \mathrm{Unif}(0.5,15),\qquad
X_{i,2}\sim \mathrm{Binom}(1,0.5),
$$

and five independent nuisance covariates

$$
Z_{i,j}\sim N(0,1),\qquad j=1,\ldots,5.
$$

The data are generated from a hurdle negative binomial model. The positive-count component has mean

$$
\log(\mu_i)=\alpha_0+\alpha_1\log(X_{i,1})+\alpha_2X_{i,2},
$$

and the hurdle, or zero-probability, component is

$$
\mathrm{logit}(\pi_i^0)=\beta_0+\beta_1X_{i,1}+\beta_2X_{i,2}.
$$

The executable settings used in this demo are

$$
\alpha_0=-1,\qquad \alpha_1=6,\qquad \alpha_2=1,
$$

$$
\beta_0=-5,\qquad \beta_1=0.6,\qquad \beta_2=-5,
$$

with negative-binomial size parameter

$$
\mathrm{size}=4.
$$

The nuisance $Z$ variables are generated with true coefficient zero and are included in fitted models to match the simulation-code structure.

## Resource and Fitting Utilities

The code below defines helper functions for reading the fixed dataset and saved `brms` fits from `resources`. It also keeps a small refitting helper for holdout fits if those fits are not included in the resource file. The Z-residual, plotting, PPC, and HPC diagnostics later in the demo are called directly from `Zresidual`.

```{r simulation-fitting-utilities}
fit_or_read_brms <- function(
  file,
  formula,
  data,
  seed,
  family,
  prior,
  chains = brms_chains,
  iter = brms_iter,
  warmup = brms_warmup
) {
  if (!force_refit && file.exists(file)) {
    return(readRDS(file))
  }

  fit <- brms::brm(
    formula = formula,
    family = family,
    prior = prior,
    data = data,
    backend = brms_backend,
    refresh = 0,
    cores = brms_cores,
    chains = chains,
    iter = iter,
    warmup = warmup,
    seed = seed,
    control = list(adapt_delta = 0.95, max_treedepth = 12),
    save_pars = brms::save_pars(all = TRUE)
  )

  saveRDS(fit, file = file)
  fit
}

read_resource_rds <- function(file, label = "resource") {
  if (!file.exists(file)) {
    stop("Cannot find ", label, ": ", file, call. = FALSE)
  }
  readRDS(file)
}

normalize_resource_name <- function(x) {
  gsub("[^a-z0-9]+", "", tolower(x))
}

list_resource_names <- function(x, prefix = character()) {
  if (!is.list(x) || inherits(x, "brmsfit")) return(character())

  nms <- names(x)
  if (is.null(nms)) nms <- rep("", length(x))

  out <- character()
  for (i in seq_along(x)) {
    this_name <- if (nzchar(nms[i])) nms[i] else paste0("[[", i, "]]")
    this_path <- paste(c(prefix, this_name), collapse = "$")
    out <- c(out, this_path)

    if (is.list(x[[i]]) && !inherits(x[[i]], "brmsfit")) {
      out <- c(out, list_resource_names(x[[i]], c(prefix, this_name)))
    }
  }

  out
}

get_resource_object <- function(x, candidates, label, required = TRUE) {
  candidates_norm <- normalize_resource_name(candidates)

  search_one <- function(obj) {
    if (!is.list(obj) || inherits(obj, "brmsfit")) return(NULL)

    nms <- names(obj)
    if (!is.null(nms)) {
      nms_norm <- normalize_resource_name(nms)
      idx <- match(candidates_norm, nms_norm, nomatch = 0L)
      idx <- idx[idx > 0L]
      if (length(idx)) return(obj[[idx[1L]]])
    }

    for (ii in seq_along(obj)) {
      if (is.list(obj[[ii]]) && !inherits(obj[[ii]], "brmsfit")) {
        ans <- search_one(obj[[ii]])
        if (!is.null(ans)) return(ans)
      }
    }

    NULL
  }

  ans <- search_one(x)

  if (is.null(ans) && isTRUE(required)) {
    available <- list_resource_names(x)
    stop(
      "Could not find ", label, " in the saved resource object.
",
      "Tried names: ", paste(candidates, collapse = ", "), "
",
      "Available names include: ", paste(utils::head(available, 40), collapse = ", "),
      call. = FALSE
    )
  }

  ans
}

get_resource_fit <- function(x, candidates, label, required = TRUE) {
  ans <- get_resource_object(x, candidates = candidates, label = label, required = required)

  if (!is.null(ans) && !inherits(ans, "brmsfit")) {
    stop(
      label, " was found, but it is not a `brmsfit` object. ",
      "Please check the saved fits file.",
      call. = FALSE
    )
  }

  ans
}
```


```{r zresidual-chisq-calculation}
zresid_chisq_pvalue <- function(zres, df) {
  if (is.list(zres) && !is.null(zres$zresidual)) {
    zres <- zres$zresidual
  }

  zmat <- as.matrix(zres)
  storage.mode(zmat) <- "numeric"

  pvals <- apply(
    zmat,
    2,
    function(z) stats::pchisq(sum(z^2, na.rm = TRUE), df = df, lower.tail = FALSE)
  )

  Zresidual::upper_bound_pvalue(pvals)
}
```

## Load the Fixed Paper Dataset

```{r load-representative-data}
sim_id <- representative_id

dat_hnb <- read_resource_rds(data_file, label = "fixed paper dataset")
dat_hnb <- tibble::as_tibble(dat_hnb)

# Keep these derived columns available even if the saved dataset only stores the minimal variables.
if (!"log_x1" %in% names(dat_hnb)) {
  dat_hnb <- dplyr::mutate(dat_hnb, log_x1 = log(x1))
}

if (!"zeros" %in% names(dat_hnb)) {
  dat_hnb <- dplyr::mutate(dat_hnb, zeros = as.integer(y == 0L))
}

missing_required <- setdiff(required_cols, names(dat_hnb))
if (length(missing_required)) {
  stop(
    "The fixed paper dataset is missing required columns: ",
    paste(missing_required, collapse = ", "),
    call. = FALSE
  )
}

zero_count_group <- factor(
  ifelse(dat_hnb$y == 0L, "zero count", "positive count"),
  levels = c("positive count", "zero count")
)

k_anova_full <- make_k_anova(nrow(dat_hnb))

dat_hnb %>%
  dplyr::summarise(
    n = dplyr::n(),
    zero_proportion = mean(zeros == 1),
    mean_y = mean(y),
    median_y = stats::median(y),
    max_y = max(y),
    k_anova = k_anova_full
  ) %>%
  gt::gt() %>%
  gt::fmt_number(columns = dplyr::everything(), decimals = 3) %>%
  gt::tab_header(title = "Fixed HNB-generated dataset summary")
```

## Fit HNB-log and HNB-linear Models

Both HNB models have the same hurdle component. They differ only in the count-component functional form of `x1`.

```{r fit-hnb-models}
#| code-fold: true
#| message: false
#| warning: false

hnb_prior <- c(
  brms::prior(normal(0, 10), class = "Intercept"),
  brms::prior(normal(0, 10), class = "b"),
  brms::prior(normal(0, 10), class = "Intercept", dpar = "hu"),
  brms::prior(normal(0, 10), class = "b", dpar = "hu"),
  brms::prior(exponential(1), class = "shape")
)

formula_log_hnb <- brms::bf(
  y ~ log_x1 + x2 + z1 + z2 + z3 + z4 + z5,
  hu ~ x1 + x2 + z1 + z2 + z3 + z4 + z5
)

formula_linear_hnb <- brms::bf(
  y ~ x1 + x2 + z1 + z2 + z3 + z4 + z5,
  hu ~ x1 + x2 + z1 + z2 + z3 + z4 + z5
)

fits_full_resource <- read_resource_rds(
  fits_full_file,
  label = "saved full-data model fits"
)

fit_log_hnb <- get_resource_fit(
  fits_full_resource,
  candidates = c("fit_log_hnb", "log_hnb", "hnb_log", "HNB_log", "HNB-log", "fit_hnb_log"),
  label = "HNB-log full-data fit"
)

fit_linear_hnb <- get_resource_fit(
  fits_full_resource,
  candidates = c("fit_linear_hnb", "linear_hnb", "hnb_linear", "HNB_linear", "HNB-linear", "fit_hnb_linear"),
  label = "HNB-linear full-data fit"
)
```

## Fit NB-log and NB-linear Comparison Models

These two comparison models are fitted to the same HNB-generated dataset. The NB-log model keeps the logarithmic functional form for `x1` but uses the wrong response family. The NB-linear model uses the wrong response family and the wrong functional form.

```{r fit-nb-models}
#| code-fold: true
#| message: false
#| warning: false

nb_prior <- c(
  brms::prior(normal(0, 10), class = "Intercept"),
  brms::prior(normal(0, 10), class = "b"),
  brms::prior(exponential(1), class = "shape")
)

formula_log_nb <- brms::bf(
  y ~ log_x1 + x2 + z1 + z2 + z3 + z4 + z5
)

formula_linear_nb <- brms::bf(
  y ~ x1 + x2 + z1 + z2 + z3 + z4 + z5
)

fit_log_nb <- get_resource_fit(
  fits_full_resource,
  candidates = c("fit_log_nb", "log_nb", "nb_log", "NB_log", "NB-log", "fit_nb_log"),
  label = "NB-log full-data fit"
)

fit_linear_nb <- get_resource_fit(
  fits_full_resource,
  candidates = c("fit_linear_nb", "linear_nb", "nb_linear", "NB_linear", "NB-linear", "fit_nb_linear"),
  label = "NB-linear full-data fit"
)
```

## Holdout Fits for PPC/HPC Comparisons

The PPC checks use the full-data fits. The HPC checks use a deterministic 20% holdout split and refit each candidate model on the training subset. This gives a compact comparison between residual-based diagnostics and predictive-check diagnostics for the same four candidate models.

```{r fit-hpc-models}
#| code-fold: true
#| message: false
#| warning: false

fits_hpc_resource <- if (file.exists(fits_hpc_file)) {
  read_resource_rds(fits_hpc_file, label = "saved holdout model fits")
} else {
  fits_full_resource
}

hpc_split <- get_resource_object(
  fits_hpc_resource,
  candidates = c("hpc_split", "holdout_split", "split_hpc", "data_split"),
  label = "HPC holdout split",
  required = FALSE
)

if (!is.null(hpc_split)) {
  dat_hnb_train <- hpc_split$train
  dat_hnb_test  <- hpc_split$test
} else {
  set.seed(seed_use + 50000 + sim_id)
  n_test <- max(1L, floor(test_prop * nrow(dat_hnb)))
  test_id <- sort(sample(seq_len(nrow(dat_hnb)), size = n_test))
  dat_hnb_train <- dat_hnb[-test_id, , drop = FALSE]
  dat_hnb_test  <- dat_hnb[ test_id, , drop = FALSE]
}

fit_log_hnb_hpc <- get_resource_fit(
  fits_hpc_resource,
  candidates = c("fit_log_hnb_hpc", "log_hnb_hpc", "hnb_log_hpc", "HNB_log_hpc", "HNB-log-hpc"),
  label = "HNB-log holdout fit",
  required = FALSE
)

if (is.null(fit_log_hnb_hpc)) {
  fit_log_hnb_hpc <- fit_or_read_brms(
    file = file.path(resources_path, paste0("fit_log_hnb_hpc", file_tag, "_sim", sim_id, ".rds")),
    formula = formula_log_hnb,
    data = dat_hnb_train,
    seed = seed_use + 60000 + sim_id,
    family = hnb_family,
    prior = hnb_prior
  )
}

fit_linear_hnb_hpc <- get_resource_fit(
  fits_hpc_resource,
  candidates = c("fit_linear_hnb_hpc", "linear_hnb_hpc", "hnb_linear_hpc", "HNB_linear_hpc", "HNB-linear-hpc"),
  label = "HNB-linear holdout fit",
  required = FALSE
)

if (is.null(fit_linear_hnb_hpc)) {
  fit_linear_hnb_hpc <- fit_or_read_brms(
    file = file.path(resources_path, paste0("fit_linear_hnb_hpc", file_tag, "_sim", sim_id, ".rds")),
    formula = formula_linear_hnb,
    data = dat_hnb_train,
    seed = seed_use + 70000 + sim_id,
    family = hnb_family,
    prior = hnb_prior
  )
}

fit_log_nb_hpc <- get_resource_fit(
  fits_hpc_resource,
  candidates = c("fit_log_nb_hpc", "log_nb_hpc", "nb_log_hpc", "NB_log_hpc", "NB-log-hpc"),
  label = "NB-log holdout fit",
  required = FALSE
)

if (is.null(fit_log_nb_hpc)) {
  fit_log_nb_hpc <- fit_or_read_brms(
    file = file.path(resources_path, paste0("fit_log_nb_hpc", file_tag, "_sim", sim_id, ".rds")),
    formula = formula_log_nb,
    data = dat_hnb_train,
    seed = seed_use + 75000 + sim_id,
    family = nb_family,
    prior = nb_prior
  )
}

fit_linear_nb_hpc <- get_resource_fit(
  fits_hpc_resource,
  candidates = c("fit_linear_nb_hpc", "linear_nb_hpc", "nb_linear_hpc", "NB_linear_hpc", "NB-linear-hpc"),
  label = "NB-linear holdout fit",
  required = FALSE
)

if (is.null(fit_linear_nb_hpc)) {
  fit_linear_nb_hpc <- fit_or_read_brms(
    file = file.path(resources_path, paste0("fit_linear_nb_hpc", file_tag, "_sim", sim_id, ".rds")),
    formula = formula_linear_nb,
    data = dat_hnb_train,
    seed = seed_use + 76000 + sim_id,
    family = nb_family,
    prior = nb_prior
  )
}
```

## Compute Posterior and ISCV RSP Z-residuals

The response is discrete, so RSP Z-residuals depend on auxiliary randomization. Here `nrep_z = 500` means 500 auxiliary randomizations for this one fixed dataset, not 500 simulated datasets.

```{r compute-zresiduals}
#| message: false
#| warning: false

zres_post_file <- file.path(
  resources_path,
  paste0("zres_post_list_all_", sample_tag, ".rds")
)

zres_iscv_file <- file.path(
  resources_path,
  paste0("zres_iscv_list_all_", sample_tag, ".rds")
)

if (
  !force_recompute_zres &&
    file.exists(zres_post_file) &&
    file.exists(zres_iscv_file)
) {
  zres_post_list <- readRDS(zres_post_file)
  zres_iscv_list <- readRDS(zres_iscv_file)

} else {
  z_hnb_log_post <- Zresidual::Zresidual(
    fit_log_hnb,
    data = dat_hnb,
    mcmc_summarize = "post",
    nrep = nrep_z
  )

  z_hnb_linear_post <- Zresidual::Zresidual(
    fit_linear_hnb,
    data = dat_hnb,
    mcmc_summarize = "post",
    nrep = nrep_z
  )

  z_nb_log_post <- Zresidual::Zresidual(
    fit_log_nb,
    data = dat_hnb,
    mcmc_summarize = "post",
    nrep = nrep_z
  )

  z_nb_linear_post <- Zresidual::Zresidual(
    fit_linear_nb,
    data = dat_hnb,
    mcmc_summarize = "post",
    nrep = nrep_z
  )

  z_hnb_log_iscv <- Zresidual::Zresidual(
    fit_log_hnb,
    data = dat_hnb,
    mcmc_summarize = "iscv",
    nrep = nrep_z
  )

  z_hnb_linear_iscv <- Zresidual::Zresidual(
    fit_linear_hnb,
    data = dat_hnb,
    mcmc_summarize = "iscv",
    nrep = nrep_z
  )

  z_nb_log_iscv <- Zresidual::Zresidual(
    fit_log_nb,
    data = dat_hnb,
    mcmc_summarize = "iscv",
    nrep = nrep_z
  )

  z_nb_linear_iscv <- Zresidual::Zresidual(
    fit_linear_nb,
    data = dat_hnb,
    mcmc_summarize = "iscv",
    nrep = nrep_z
  )

  zres_post_list <- list(
    HNB_log = z_hnb_log_post,
    HNB_linear = z_hnb_linear_post,
    NB_log = z_nb_log_post,
    NB_linear = z_nb_linear_post
  )

  zres_iscv_list <- list(
    HNB_log = z_hnb_log_iscv,
    HNB_linear = z_hnb_linear_iscv,
    NB_log = z_nb_log_iscv,
    NB_linear = z_nb_linear_iscv
  )

  saveRDS(zres_post_list, zres_post_file)
  saveRDS(zres_iscv_list, zres_iscv_file)
}

# Unpack named lists back to the object names used later in the vignette
z_hnb_log_post    <- zres_post_list$HNB_log
z_hnb_linear_post <- zres_post_list$HNB_linear
z_nb_log_post     <- zres_post_list$NB_log
z_nb_linear_post  <- zres_post_list$NB_linear

z_hnb_log_iscv    <- zres_iscv_list$HNB_log
z_hnb_linear_iscv <- zres_iscv_list$HNB_linear
z_nb_log_iscv     <- zres_iscv_list$NB_log
z_nb_linear_iscv  <- zres_iscv_list$NB_linear

zcov_hnb_log    <- Zresidual::Zcov(fit_log_hnb,    data = dat_hnb)
zcov_hnb_linear <- Zresidual::Zcov(fit_linear_hnb, data = dat_hnb)
zcov_nb_log     <- Zresidual::Zcov(fit_log_nb,     data = dat_hnb)
zcov_nb_linear  <- Zresidual::Zcov(fit_linear_nb,  data = dat_hnb)

fits_full <- list(
  HNB_log = fit_log_hnb,
  HNB_linear = fit_linear_hnb,
  NB_log = fit_log_nb,
  NB_linear = fit_linear_nb
)

p_eff <- vapply(
  fits_full,
  function(fit) loo::loo(fit)$estimates["p_loo", "Estimate"],
  numeric(1L)
)

df_post <- vapply(
  names(zres_post_list),
  function(mm) max(1, nrow(as.matrix(zres_post_list[[mm]])) - p_eff[[mm]]),
  numeric(1L)
)

df_iscv <- vapply(
  zres_iscv_list,
  function(zres) nrow(as.matrix(zres)),
  numeric(1L)
)

# Plot-only helper: force zero/count display without changing package functions
add_zero_indicator_for_plot <- function(zres, y) {
  z_plot <- zres
  attr(z_plot, "type") <- "hurdle"
  attr(z_plot, "zero_id") <- which(y == 0L)
  z_plot
}

z_hnb_log_post_plot    <- add_zero_indicator_for_plot(z_hnb_log_post, dat_hnb$y)
z_hnb_linear_post_plot <- add_zero_indicator_for_plot(z_hnb_linear_post, dat_hnb$y)
z_nb_log_post_plot     <- add_zero_indicator_for_plot(z_nb_log_post, dat_hnb$y)
z_nb_linear_post_plot  <- add_zero_indicator_for_plot(z_nb_linear_post, dat_hnb$y)

z_hnb_log_iscv_plot    <- add_zero_indicator_for_plot(z_hnb_log_iscv, dat_hnb$y)
z_hnb_linear_iscv_plot <- add_zero_indicator_for_plot(z_hnb_linear_iscv, dat_hnb$y)
z_nb_log_iscv_plot     <- add_zero_indicator_for_plot(z_nb_log_iscv, dat_hnb$y)
z_nb_linear_iscv_plot  <- add_zero_indicator_for_plot(z_nb_linear_iscv, dat_hnb$y)

```

## Visual Z-residual Diagnostics for the HNB Models

The first diagnostic view compares posterior RSP Z-residuals against `x1` for the HNB-log and HNB-linear models. The HNB-log model is the data-generating model and should not show a systematic residual mean pattern. The HNB-linear model keeps the correct response family but uses a wrong linear functional form, so a residual pattern against `x1` is expected. LOWESS curves are omitted because the visual and numerical conclusions rely on the residual points, binned boxplots, and diagnostic p-values rather than on a smoothing algorithm.

```{r fig-hnb-post-scatter-boxplot}
#| fig-cap: "Posterior RSP Z-residual diagnostics for the HNB-log and HNB-linear models fitted to the same fixed HNB-generated dataset. Scatterplot points are grouped by the observed zero-count indicator."
#| fig-width: 18
#| fig-height: 4.6
#| out-width: 100%
#| dpi: 300
#| message: false
#| results: false

irep_plot <- 1

op <- par(no.readonly = TRUE)
par(mfrow = c(1, 4), mar = c(4, 4, 3, 1), oma = c(0, 0, 1, 0), cex = 1.02)

plot(
  z_hnb_log_post_plot,
  zcov = zcov_hnb_log,
  x_axis_var = "x1",
  irep = irep_plot,
  add_lowess = TRUE,
  main.title = expression("(a) HNB Log, " * Z^"Post-RSP" * " Scatterplot")
)

boxplot(
  z_hnb_log_post,
  zcov = zcov_hnb_log,
  x_axis_var = "x1",
  irep = irep_plot,
  num.bin = k_anova_full,
  k.test = k_anova_full,
  main.title = expression("(b) HNB Log, " * Z^"Post-RSP" * " Boxplot")
)

plot(
  z_hnb_linear_post_plot,
  zcov = zcov_hnb_linear,
  x_axis_var = "x1",
  irep = irep_plot,
  add_lowess = TRUE,
  main.title = expression("(c) HNB Linear, " * Z^"Post-RSP" * " Scatterplot")
)

boxplot(
  z_hnb_linear_post,
  zcov = zcov_hnb_linear,
  x_axis_var = "x1",
  irep = irep_plot,
  num.bin = k_anova_full,
  k.test = k_anova_full,
  main.title = expression("(d) HNB Linear, " * Z^"Post-RSP" * " Boxplot")
)

par(op)
```

## Visual Z-residual Diagnostics for the NB Comparison Models

The second diagnostic view shows the NB-log and NB-linear comparison models. For the NB-log model, the response family is misspecified because the data were generated from a hurdle negative binomial process. In the HNB data-generating process, the probability of entering the count component changes with the covariate. When a standard NB model is fitted without an explicit hurdle component, this zero/count mechanism can be projected into the NB mean structure. Therefore, an AOV rejection against `x1` for the NB-log model is informative and should not be interpreted as a contradiction. Points in the scatterplots are grouped by the observed zero-count indicator for both NB models, because separation or over-abundance of small residuals among zero observations is informative about errors in the zero-count mechanism.

```{r fig-nb-post-scatter-boxplot}
#| fig-cap: "Posterior RSP Z-residual diagnostics for the NB-log and NB-linear comparison models fitted to the fixed HNB-generated dataset. Scatterplot points are grouped by the observed zero-count indicator."
#| fig-width: 18
#| fig-height: 4.6
#| out-width: 100%
#| dpi: 300
#| message: false
#| results: false

irep_plot <- 1

op <- par(no.readonly = TRUE)
par(mfrow = c(1, 4), mar = c(4, 4, 3, 1), oma = c(0, 0, 1, 0), cex = 1.02)

plot(
  z_nb_log_post_plot,
  zcov = zcov_nb_log,
  x_axis_var = "x1",
  irep = irep_plot,
  add_lowess = TRUE,
  main.title = expression("(a) NB Log, " * Z^"Post-RSP" * " Scatterplot")
)

boxplot(
  z_nb_log_post,
  zcov = zcov_nb_log,
  x_axis_var = "x1",
  irep = irep_plot,
  num.bin = k_anova_full,
  k.test = k_anova_full,
  main.title = expression("(b) NB Log, " * Z^"Post-RSP" * " Boxplot")
)

plot(
  z_nb_linear_post_plot,
  zcov = zcov_nb_linear,
  x_axis_var = "x1",
  irep = irep_plot,
  add_lowess = TRUE,
  main.title = expression("(c) NB Linear, " * Z^"Post-RSP" * " Scatterplot")
)

boxplot(
  z_nb_linear_post,
  zcov = zcov_nb_linear,
  x_axis_var = "x1",
  irep = irep_plot,
  num.bin = k_anova_full,
  k.test = k_anova_full,
  main.title = expression("(d) NB Linear, " * Z^"Post-RSP" * " Boxplot")
)

par(op)
```

## PPC and HPC Checks

```{r compute-ppc-hpc}
#| code-fold: true
#| message: false
#| warning: false

ppc_file <- file.path(
  resources_path,
  paste0("ppc_list_all_", sample_tag, ".rds")
)

hpc_file <- file.path(
  resources_path,
  paste0("hpc_list_all_", sample_tag, ".rds")
)

if (
  !force_recompute_ppc_hpc &&
    file.exists(ppc_file) &&
    file.exists(hpc_file)
) {
  ppc_list <- readRDS(ppc_file)
  hpc_list <- readRDS(hpc_file)

} else {
  ppc_list <- list(
    HNB_log = Zresidual::ppc(
      fit = fit_log_hnb,
      data = dat_hnb,
      x = x_check,
      ndraws = ndraws_use,
      seed = seed_use + 80000 + sim_id,
      randomized = TRUE
    ),

    HNB_linear = Zresidual::ppc(
      fit = fit_linear_hnb,
      data = dat_hnb,
      x = x_check,
      ndraws = ndraws_use,
      seed = seed_use + 80001 + sim_id,
      randomized = TRUE
    ),

    NB_log = Zresidual::ppc(
      fit = fit_log_nb,
      data = dat_hnb,
      x = x_check,
      ndraws = ndraws_use,
      seed = seed_use + 80002 + sim_id,
      randomized = TRUE
    ),

    NB_linear = Zresidual::ppc(
      fit = fit_linear_nb,
      data = dat_hnb,
      x = x_check,
      ndraws = ndraws_use,
      seed = seed_use + 80003 + sim_id,
      randomized = TRUE
    )
  )

  hpc_list <- list(
    HNB_log = Zresidual::hpc(
      fit = fit_log_hnb_hpc,
      newdata = dat_hnb_test,
      x = x_check,
      ndraws = ndraws_use,
      seed = seed_use + 90000 + sim_id,
      k_anova = 5,
      k_bl = 5,
      randomized = TRUE
    ),

    HNB_linear = Zresidual::hpc(
      fit = fit_linear_hnb_hpc,
      newdata = dat_hnb_test,
      x = x_check,
      ndraws = ndraws_use,
      seed = seed_use + 90001 + sim_id,
      k_anova = 5,
      k_bl = 5,
      randomized = TRUE
    ),

    NB_log = Zresidual::hpc(
      fit = fit_log_nb_hpc,
      newdata = dat_hnb_test,
      x = x_check,
      ndraws = ndraws_use,
      seed = seed_use + 90002 + sim_id,
      k_anova = 5,
      k_bl = 5,
      randomized = TRUE
    ),

    NB_linear = Zresidual::hpc(
      fit = fit_linear_nb_hpc,
      newdata = dat_hnb_test,
      x = x_check,
      ndraws = ndraws_use,
      seed = seed_use + 90003 + sim_id,
      k_anova = 5,
      k_bl = 5,
      randomized = TRUE
    )
  )

  saveRDS(ppc_list, ppc_file)
  saveRDS(hpc_list, hpc_file)
}

# Unpack back to the names used later in the table
ppc_hnb_log    <- ppc_list$HNB_log
ppc_hnb_linear <- ppc_list$HNB_linear
ppc_nb_log     <- ppc_list$NB_log
ppc_nb_linear  <- ppc_list$NB_linear

hpc_hnb_log    <- hpc_list$HNB_log
hpc_hnb_linear <- hpc_list$HNB_linear
hpc_nb_log     <- hpc_list$NB_log
hpc_nb_linear  <- hpc_list$NB_linear
```

## Replicated Z-residual p-value Histograms

For discrete responses, RSP Z-residuals depend on the auxiliary randomization used to smooth the probability mass at each observed count. Therefore, the same fitted model can produce a distribution of diagnostic p-values across repeated auxiliary randomizations. Here we use `nrep_z = 500` replicated Z-residual p-values for each model and show two diagnostics: the overall Z-SW test and the targeted Z-AOV test against `x1`.

The vertical line in each histogram marks the distribution-free upper-bound p-value used as the single-number summary in the table below. The HNB-log model is the true model. The HNB-linear model represents a wrong functional form under the correct HNB response family. The NB-log and NB-linear models illustrate how response-family misspecification can also induce apparent covariate-specific residual structure when the omitted hurdle zero/count mechanism is projected into the NB mean structure.

```{r replicated-z-pvalue-histograms}
#| fig-cap: "Replicated posterior RSP Z-residual p-values for Z-SW and Z-AOV-X1 across one true model and three misspecified models. Each histogram is based on 500 auxiliary randomizations."
#| fig-width: 16
#| fig-height: 7.8
#| out-width: 100%
#| dpi: 300
#| message: false
#| results: false

post_sw_pvals <- list(
  `HNB-log` = sw.test.zresid(z_hnb_log_post),
  `HNB-linear` = sw.test.zresid(z_hnb_linear_post),
  `NB-log` = sw.test.zresid(z_nb_log_post),
  `NB-linear` = sw.test.zresid(z_nb_linear_post)
)

post_aov_x1_pvals <- list(
  `HNB-log` = aov.test.zresid(z_hnb_log_post, X = "x1", zcov = zcov_hnb_log, k_anova = k_anova_full),
  `HNB-linear` = aov.test.zresid(z_hnb_linear_post, X = "x1", zcov = zcov_hnb_linear, k_anova = k_anova_full),
  `NB-log` = aov.test.zresid(z_nb_log_post, X = "x1", zcov = zcov_nb_log, k_anova = k_anova_full),
  `NB-linear` = aov.test.zresid(z_nb_linear_post, X = "x1", zcov = zcov_nb_linear, k_anova = k_anova_full)
)

post_sw_ub <- vapply(post_sw_pvals, Zresidual::upper_bound_pvalue, numeric(1L))
post_aov_x1_ub <- vapply(post_aov_x1_pvals, Zresidual::upper_bound_pvalue, numeric(1L))

plot_hist_pvalues <- function(pvals, ub, main) {
  pvals <- pvals[is.finite(pvals)]
  hist(
    pvals,
    breaks = seq(0, 1, by = 0.05),
    xlim = c(0, 1),
    main = main,
    xlab = "Replicated p-values",
    ylab = "Frequency",
    col = "grey85",
    border = "white"
  )
  abline(v = ub, col = "red", lwd = 2)
  abline(v = 0.05, lty = 2, lwd = 1)
  legend(
    "topright",
    legend = paste0("UB = ", sprintf("%.3f", ub)),
    bty = "n",
    cex = 0.85
  )
}

op <- par(no.readonly = TRUE)
par(mfrow = c(2, 4), mar = c(4, 4, 3, 1), oma = c(0, 0, 1, 0), cex = 1.0)

plot_hist_pvalues(post_sw_pvals[["HNB-log"]], post_sw_ub[["HNB-log"]], "(a) HNB-log: Z-SW")
plot_hist_pvalues(post_sw_pvals[["HNB-linear"]], post_sw_ub[["HNB-linear"]], "(b) HNB-linear: Z-SW")
plot_hist_pvalues(post_sw_pvals[["NB-log"]], post_sw_ub[["NB-log"]], "(c) NB-log: Z-SW")
plot_hist_pvalues(post_sw_pvals[["NB-linear"]], post_sw_ub[["NB-linear"]], "(d) NB-linear: Z-SW")

plot_hist_pvalues(post_aov_x1_pvals[["HNB-log"]], post_aov_x1_ub[["HNB-log"]], "(e) HNB-log: Z-AOV-X1")
plot_hist_pvalues(post_aov_x1_pvals[["HNB-linear"]], post_aov_x1_ub[["HNB-linear"]], "(f) HNB-linear: Z-AOV-X1")
plot_hist_pvalues(post_aov_x1_pvals[["NB-log"]], post_aov_x1_ub[["NB-log"]], "(g) NB-log: Z-AOV-X1")
plot_hist_pvalues(post_aov_x1_pvals[["NB-linear"]], post_aov_x1_ub[["NB-linear"]], "(h) NB-linear: Z-AOV-X1")

par(op)
```

## Diagnostic P-value Summary

The table reports the same one-dataset diagnostic workflow for all four models. For posterior and ISCV Z-residual tests, the entries are upper-bound p-values summarizing 500 replicated RSP Z-residual p-values. PPC and HPC entries are the corresponding predictive-check p-values. The most important targeted diagnostic here is Z-AOV-X1, because the deliberate functional-form error is tied to the effect of `x1`.

```{r diagnostic-results-table}
#| code-fold: true

iscv_sw_pvals <- list(
  `HNB-log` = sw.test.zresid(z_hnb_log_iscv),
  `HNB-linear` = sw.test.zresid(z_hnb_linear_iscv),
  `NB-log` = sw.test.zresid(z_nb_log_iscv),
  `NB-linear` = sw.test.zresid(z_nb_linear_iscv)
)

iscv_aov_x1_pvals <- list(
  `HNB-log` = aov.test.zresid(z_hnb_log_iscv, X = "x1", zcov = zcov_hnb_log, k_anova = k_anova_full),
  `HNB-linear` = aov.test.zresid(z_hnb_linear_iscv, X = "x1", zcov = zcov_hnb_linear, k_anova = k_anova_full),
  `NB-log` = aov.test.zresid(z_nb_log_iscv, X = "x1", zcov = zcov_nb_log, k_anova = k_anova_full),
  `NB-linear` = aov.test.zresid(z_nb_linear_iscv, X = "x1", zcov = zcov_nb_linear, k_anova = k_anova_full)
)

post_aov_lp_pvals <- list(
  `HNB-log` = aov.test.zresid(z_hnb_log_post, X = "lp", zcov = zcov_hnb_log, k_anova = k_anova_full),
  `HNB-linear` = aov.test.zresid(z_hnb_linear_post, X = "lp", zcov = zcov_hnb_linear, k_anova = k_anova_full),
  `NB-log` = aov.test.zresid(z_nb_log_post, X = "lp", zcov = zcov_nb_log, k_anova = k_anova_full),
  `NB-linear` = aov.test.zresid(z_nb_linear_post, X = "lp", zcov = zcov_nb_linear, k_anova = k_anova_full)
)

post_bl_x1_pvals <- list(
  `HNB-log` = bartlett.test.zresid(z_hnb_log_post, X = "x1", zcov = zcov_hnb_log, k_anova = k_anova_full),
  `HNB-linear` = bartlett.test.zresid(z_hnb_linear_post, X = "x1", zcov = zcov_hnb_linear, k_anova = k_anova_full),
  `NB-log` = bartlett.test.zresid(z_nb_log_post, X = "x1", zcov = zcov_nb_log, k_anova = k_anova_full),
  `NB-linear` = bartlett.test.zresid(z_nb_linear_post, X = "x1", zcov = zcov_nb_linear, k_anova = k_anova_full)
)

iscv_aov_lp_pvals <- list(
  `HNB-log` = aov.test.zresid(z_hnb_log_iscv, X = "lp", zcov = zcov_hnb_log, k_anova = k_anova_full),
  `HNB-linear` = aov.test.zresid(z_hnb_linear_iscv, X = "lp", zcov = zcov_hnb_linear, k_anova = k_anova_full),
  `NB-log` = aov.test.zresid(z_nb_log_iscv, X = "lp", zcov = zcov_nb_log, k_anova = k_anova_full),
  `NB-linear` = aov.test.zresid(z_nb_linear_iscv, X = "lp", zcov = zcov_nb_linear, k_anova = k_anova_full)
)

iscv_bl_x1_pvals <- list(
  `HNB-log` = bartlett.test.zresid(z_hnb_log_iscv, X = "x1", zcov = zcov_hnb_log, k_anova = k_anova_full),
  `HNB-linear` = bartlett.test.zresid(z_hnb_linear_iscv, X = "x1", zcov = zcov_hnb_linear, k_anova = k_anova_full),
  `NB-log` = bartlett.test.zresid(z_nb_log_iscv, X = "x1", zcov = zcov_nb_log, k_anova = k_anova_full),
  `NB-linear` = bartlett.test.zresid(z_nb_linear_iscv, X = "x1", zcov = zcov_nb_linear, k_anova = k_anova_full)
)

model_order <- c("HNB-log", "HNB-linear", "NB-log", "NB-linear")

summary_table <- tibble::tibble(
  Model = model_order,

  Post_SW = post_sw_ub[model_order],
  Post_AOV_x1 = post_aov_x1_ub[model_order],
  Post_AOV_LP = vapply(post_aov_lp_pvals[model_order], Zresidual::upper_bound_pvalue, numeric(1L)),
  Post_BL_x1 = vapply(post_bl_x1_pvals[model_order], Zresidual::upper_bound_pvalue, numeric(1L)),
  Post_chisq = c(
    zresid_chisq_pvalue(z_hnb_log_post, df = df_post[["HNB_log"]]),
    zresid_chisq_pvalue(z_hnb_linear_post, df = df_post[["HNB_linear"]]),
    zresid_chisq_pvalue(z_nb_log_post, df = df_post[["NB_log"]]),
    zresid_chisq_pvalue(z_nb_linear_post, df = df_post[["NB_linear"]])
  ),

  ISCV_SW = vapply(iscv_sw_pvals[model_order], Zresidual::upper_bound_pvalue, numeric(1L)),
  ISCV_AOV_x1 = vapply(iscv_aov_x1_pvals[model_order], Zresidual::upper_bound_pvalue, numeric(1L)),
  ISCV_AOV_LP = vapply(iscv_aov_lp_pvals[model_order], Zresidual::upper_bound_pvalue, numeric(1L)),
  ISCV_BL_x1 = vapply(iscv_bl_x1_pvals[model_order], Zresidual::upper_bound_pvalue, numeric(1L)),
  ISCV_chisq = c(
    zresid_chisq_pvalue(z_hnb_log_iscv, df = df_iscv[["HNB_log"]]),
    zresid_chisq_pvalue(z_hnb_linear_iscv, df = df_iscv[["HNB_linear"]]),
    zresid_chisq_pvalue(z_nb_log_iscv, df = df_iscv[["NB_log"]]),
    zresid_chisq_pvalue(z_nb_linear_iscv, df = df_iscv[["NB_linear"]])
  ),

  PPC_SW = c(ppc_hnb_log$sw_rpp, ppc_hnb_linear$sw_rpp, ppc_nb_log$sw_rpp, ppc_nb_linear$sw_rpp),
  PPC_AOV_x1 = c(
    ppc_hnb_log$aov_rpp_all[["AOV[x1]"]],
    ppc_hnb_linear$aov_rpp_all[["AOV[x1]"]],
    ppc_nb_log$aov_rpp_all[["AOV[x1]"]],
    ppc_nb_linear$aov_rpp_all[["AOV[x1]"]]
  ),
  PPC_AOV_LP = c(
    ppc_hnb_log$aov_rpp_all[["AOV[lp]"]],
    ppc_hnb_linear$aov_rpp_all[["AOV[lp]"]],
    ppc_nb_log$aov_rpp_all[["AOV[lp]"]],
    ppc_nb_linear$aov_rpp_all[["AOV[lp]"]]
  ),
  PPC_y_chisq = c(ppc_hnb_log$chisq, ppc_hnb_linear$chisq, ppc_nb_log$chisq, ppc_nb_linear$chisq),

  HPC_SW = c(hpc_hnb_log$sw_rpp, hpc_hnb_linear$sw_rpp, hpc_nb_log$sw_rpp, hpc_nb_linear$sw_rpp),
  HPC_AOV_x1 = c(
    hpc_hnb_log$aov_rpp_all[["AOV[x1]"]],
    hpc_hnb_linear$aov_rpp_all[["AOV[x1]"]],
    hpc_nb_log$aov_rpp_all[["AOV[x1]"]],
    hpc_nb_linear$aov_rpp_all[["AOV[x1]"]]
  ),
  HPC_AOV_LP = c(
    hpc_hnb_log$aov_rpp_all[["AOV[lp]"]],
    hpc_hnb_linear$aov_rpp_all[["AOV[lp]"]],
    hpc_nb_log$aov_rpp_all[["AOV[lp]"]],
    hpc_nb_linear$aov_rpp_all[["AOV[lp]"]]
  ),
  HPC_y_chisq = c(hpc_hnb_log$chisq, hpc_hnb_linear$chisq, hpc_nb_log$chisq, hpc_nb_linear$chisq)
)

gt_tbl <- summary_table %>%
  gt::gt() %>%
  gt::tab_spanner(
    label = "Posterior RSP Z-residuals",
    columns = c(Post_SW, Post_AOV_x1, Post_AOV_LP, Post_BL_x1, Post_chisq)
  ) %>%
  gt::tab_spanner(
    label = "ISCV RSP Z-residuals",
    columns = c(ISCV_SW, ISCV_AOV_x1, ISCV_AOV_LP, ISCV_BL_x1, ISCV_chisq)
  ) %>%
  gt::tab_spanner(
    label = "PPC tests",
    columns = c(PPC_SW, PPC_AOV_x1, PPC_AOV_LP, PPC_y_chisq)
  ) %>%
  gt::tab_spanner(
    label = "HPC tests",
    columns = c(HPC_SW, HPC_AOV_x1, HPC_AOV_LP, HPC_y_chisq)
  ) %>%
  gt::cols_label(
    Model = "Model",
    Post_SW = "Z-SW",
    Post_AOV_x1 = "Z-AOV: x1",
    Post_AOV_LP = "Z-AOV: LP",
    Post_BL_x1 = "Z-BL: x1",
    Post_chisq = gt::md("Z-$\\chi^2$"),
    ISCV_SW = "Z-SW",
    ISCV_AOV_x1 = "Z-AOV: x1",
    ISCV_AOV_LP = "Z-AOV: LP",
    ISCV_BL_x1 = "Z-BL: x1",
    ISCV_chisq = gt::md("Z-$\\chi^2$"),
    PPC_SW = "Z-SW",
    PPC_AOV_x1 = "Z-AOV: x1",
    PPC_AOV_LP = "Z-AOV: LP",
    PPC_y_chisq = gt::md("$y$-$\\chi^2$"),
    HPC_SW = "Z-SW",
    HPC_AOV_x1 = "Z-AOV: x1",
    HPC_AOV_LP = "Z-AOV: LP",
    HPC_y_chisq = gt::md("$y$-$\\chi^2$")
  ) %>%
  gt::fmt_number(columns = -c(Model), decimals = 3) %>%
  gt::cols_align(align = "left", columns = Model) %>%
  gt::cols_align(align = "center", columns = -c(Model)) %>%
  gt::cols_width(
    Post_SW:HPC_y_chisq ~ gt::px(78),
    Model ~ gt::px(130)
  ) %>%
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_body(columns = Model)
  ) %>%
  gt::tab_options(
    table.width = gt::px(1600),
    table.font.size = gt::px(12),
    heading.title.font.size = gt::px(16),
    column_labels.font.size = gt::px(11),
    data_row.padding = gt::px(4)
  ) %>%
  gt::tab_header(title = "Representative HNB example: diagnostic p-values") %>%
  gt::tab_source_note(
    source_note = "Posterior and ISCV Z-residual entries are distribution-free upper-bound p-values based on 500 auxiliary randomizations. PPC and HPC entries are predictive-check p-values."
  )

htmltools::div(
  style = "width: 100%; overflow-x: auto; padding-bottom: 10px;",
  htmltools::HTML(gt::as_raw_html(gt_tbl, inline_css = TRUE))
)
```

## Interpretation

The HNB-log model is the reference case because it uses both the data-generating response family and the data-generating logarithmic effect of `x1`. Its residual plots should look comparatively structureless, and the targeted Z-AOV-X1 p-values should not show systematic evidence of a residual mean shift across `x1` groups.

The HNB-linear model isolates a functional-form error. A small Z-AOV-X1 p-value for this model indicates that replacing the logarithmic effect with a linear effect leaves a systematic residual pattern against `x1`. This is the main advantage of covariate-specific Z-residual diagnostics: they can identify where the model is wrong, rather than only saying that the overall fit is inadequate.

The NB-log model does not isolate only a response-family error. Because the HNB data-generating process has a covariate-dependent zero/count mechanism, fitting an ordinary NB model can project this omitted hurdle structure into the NB mean function. Consequently, a Z-AOV-X1 rejection for the NB-log model is meaningful: it reflects the effective functional-form distortion induced by the wrong response family, not an inconsistency in the diagnostic. The zero-count indicators in the NB residual plots are useful even though the NB models do not explicitly contain a hurdle component: isolation or over-abundance of small residuals among zeros can reveal errors in modelling the distribution of zero counts. The NB-linear model combines the response-family misspecification with an explicitly wrong linear functional form and can therefore be flagged by both overall and covariate-specific diagnostics.

Taken together, the demo illustrates a practical residual workflow for Bayesian discrete regression: use Z-residual plots to see the structure of the misfit, use replicated Z-residual p-values and upper-bound summaries for discrete responses, and compare the conclusions with PPC and HPC checks. These results also suggest that different residual diagnostics may respond to different sources of model misspecification. Developing diagnostic procedures that can systematically distinguish among misspecified mean structure, zero-generating mechanism, and count distributional form is an interesting direction for future work and would require dedicated simulation studies.
